---
title: "Readme MetaGaia"
ate: '`r format(Sys.Date(),"%d/%b/%Y")`'
output:
  html_document:
    rmdformats::downcute
  toc: true
  toc_float: true
---

```{r knitr setup, include=FALSE,  eval=TRUE, echo=FALSE, warning=FALSE}
library(knitr)
knitr::opts_chunk$set(eval=TRUE, cache=FALSE, message=FALSE, warning=FALSE, 
                      comment = "", results="markup")
```

# About MetaGaia

MetaGaia is a pipeline that streamlines the process of calculating bin abundances, identifying the metabolic pathways present in each bin, and the metabolic pathways found in both phages and their hosts. The importance of performing each of these functions is to better understand microbial communities and the dominant microbes present within a community. This pipeline is divided into three sections that perform different functions: 1) Calculating bin abundances (abundance), 2) Identifying metabolic pathways present in each bin (metabolism), and 3) Finding commonalities between phages and their hosts (host-virus). Multiple requirements are needed and are specified in the "Requirements" section of this protocol.

# MetaGaia installation 

The MetaGaia software is available as an open-source package distributed from a GitHub repository. Thus,
the natural way of installing it is by cloning the repository via the following commands:
```{bash, eval=FALSE}
git clone https://github.com/valdeanda/MetaGaia
```

# Python installation requirements (Linux)

To install python, run `sudo apt-get install python3`.
To install pip, run `sudo apt install python3-pip`.
To install pandas, run `pip install pandas`.
To install numpy, run `pip install numpy`.
To install seaborn, run `pip install seaborn`.

# MetaGaia requirements

*NOTE: Please do NOT use symbolic links for fastq or fna files. Also, make sure headers are present and properly separated in input files.*

## Calculating bin abundances (abundance)

1. All necessary fastq files have been downloaded and placed within the same directory. All files with the .fastq.gz extension in a directory will be analyzed.
2. All necessary fna files have been downloaded and placed within the same directory.
3. All necessary depth text files have been downloaded and placed within the same directory or a concatenated depth file has already been created.
4. A file has been created that maps each bin to its respective sample site (see "Mapping bins to sample sites" for more information).
5. The GTDBTK tool has been downloaded to map each bin to its respective taxa (see "Mapping taxa" for more information).

## Identifying metabolic pathways (metabolism)

1. The `mapping_file.tsv` file created from the "bin_abundance" section of the pipeline.
2. An account at the Integrated Microbial Genomes with Microbiome Samples system [IMG](https://img.jgi.doe.gov/) is needed with the IMG annotation of all of the samples being analyzed.
3. File to combine annotation files. (What is this file?)
4. A user created file that contains the KEGG, COG, PFAM, and/or EC_NUMBER pathways of interest.

## Finding commonalities between phages and their hosts (host-virus)

1. The "database_counts" file from the `merge_databases.py` script that contains only one of the following databases: KEGG, COG, PFAM, or EC_NUMBER.
2. The output from Vibrant that maps phage scaffolds to metabolic pathways.

# Running abundance

## Creating bin abundance input files

1. To create all the input files necessary, run the `files_prep.py` script. Two input files (concatenated `depth` file is optional) are necessary and the directories where the `fastq` and `fna ` (and optional `depth`) files are located are needed to run this script (see "Download fastqs" for more information on where to obtain these). One of the depth arguments must be present: depth or depth_dir. These are the arguments necessary to run this script:

```{bash, eval=FALSE}
usage: files_prep.py [-h] -b BIN_SAMPLES [-d DEPTH] [-f DEPTH_DIR] -q FASTQ_DIR -a FNA_DIR

optional arguments:
  -h, --help            show this help message and exit
  -b BIN_SAMPLES, --bin_samples BIN_SAMPLES
                        Input file containing 2 columns: 1st column contains bin names and 2nd column contains sample names. Requires header for each column, "Bin" and "Sampling_Site"
                        respectively. File must have tsv, csv, or txt file extension.
  -d DEPTH, --depth DEPTH
                        Input file containing 5 columns: contigName, contigLen, totalAvgDepth, Sample_Depth, and Depth. This file can be created with jgi_summarize_bam_contig_depths. File must
                        have tsv, csv, or txt file extension.
  -f DEPTH_DIR, --depth_dir DEPTH_DIR
                        Input directory path containing all the depth txt files that must be concatenated. No other txt files should be present within the directory!
  -q FASTQ_DIR, --fastq_dir FASTQ_DIR
                        Input directory path containing all the fasta files.
  -a FNA_DIR, --fna_dir FNA_DIR
                        Input directory path containing all the fna files.
```

Example of concatenated depth file (NOTE: The Sample names in the depth file may need to be edited by the user to match the Sample names in the reads_file.tsv. WARNING: Samples without depth values are removed if present.):

```{bash, eval=FALSE}
head depth_file.tsv
```

```{r table1, echo=FALSE, message=FALSE, warnings=FALSE, results='asis'}
tabl_1 <- "

| contigName      | contigLen         | TotalAvgDepth      | SF_Jul11_USGS_4_1-sorted.bam   |
|---------------|:-------------|:-------------|:-------------:|
| S4B_scaffold_41967_c1   | 2430.0    | 15.7917   | 15.7917   |
| S4B_scaffold_22085_c1   | 3249.0    | 12.081   | 12.081   |
| S4B_scaffold_39665_c1   | 2489.0    | 9.998289999999999   | 9.998289999999999   |
| S4B_scaffold_36461_c1   | 2583.0    | 8.70695   | 8.70695   |
| S4B_scaffold_62858_c1   | 2036.0    | 11.1448   | 11.1448   |
| S4B_scaffold_47653_c1   | 2298.0    | 10.1196   | 10.1196   |
| S4B_scaffold_28262_c1   | 2900.0    | 8.95927   | 8.95927   |
"
cat(tabl_1) # output the table in a format good for HTML/PDF/docx conversion
```

To create the `bin_sample` file, read the "Mapping bins to samples" section below. To create the `depth` file, read the "Create initial depth file" section below as well.

Example of bin_sample file:

```{bash, eval=FALSE}
head bin_sample.tsv
```

```{r table2, echo=FALSE, message=FALSE, warnings=FALSE, results='asis'}
tabl2 <- "

| Bin                 | Sampling_Site
|---------------|:---------------:|
| NoBin               | 4.1   |
| 13_Jan_SF_Bin44     | 8.1   |
| 14_Jan_SF_Bin36     | 13   |
| 21_Jan_SF_Bin21     | 24   |
| 24_Jan_SF_Bin43     | 21   |
"

cat(tabl2)
```

Here is an example of how to run this script:

```{bash, eval=FALSE}
python3 files_prep.py -b ../../data/example_input_files/files_prep/bin_sample.tsv -d ../../data/example_input_files/files_prep/depth.tsv -q ~/fastq_files/ -a ~/fna_files/
```
or (if depth files are not already concatenated):
```{bash, eval=FALSE}
python3 files_prep.py -b ../../data/example_input_files/files_prep/bin_sample.tsv -f ~/depth_files/ -q ~/fastq_files/ -a ~/fna_files/
```

2. There should be four files named "reads_file.tsv" (Sample names may need to be edited by user), "mapping_file.tsv", "depth_file", and "binsize_file.tsv" saved in the `output` folder.

3. From this output, you need to rename the reads_file.tsv and depth_file.tsv. These files will look like this:
```{bash, eval = FALSE}
head reads_file.tsv
```

```{r table3, echo=FALSE, message=FALSE, warnings=FALSE, results='asis'}
tabl_3 <- "

| Sample      | Reads          | 
|---------------|:-------------|
| /home/mlangwig/Working_Dir/SF_Bay/fastq/SF_Jul11_sed_USGS_4_1_filter-METAGENOME.fastq.gz      | 135081488 |
| /home/mlangwig/Working_Dir/SF_Bay/fastq/SF_Jan12_sed_USGS_4_1_filter-METAGENOME.fastq.gz      | 140529442   |
"
cat(tabl_3) # output the table in a format good for HTML/PDF/docx conversion
```

```{bash, eval = FALSE}
head depth_file_renamed.tsv
```

```{r table4, echo=FALSE, message=FALSE, warnings=FALSE, results='asis'}
tabl_4 <- "

| Original_Contig_Name        | contigLen           | totalAvgDepth        | Sample           | Depth       | 
|---------------|:-------------|:---------------|:-------------|:---------------:| 
|S4C_scaffold_7529_c1 | 6042.0  | 23.1232  | SF_Jul11_USGS_4_1-sorted.bam   | 23.1232  |
|S4C_scaffold_8778_c1 | 5493.0  | 9.78027  | SF_Jul11_USGS_4_1-sorted.bam   | 9.78027  |
"
cat(tabl_4) # output the table in a format good for HTML/PDF/docx conversion
```

4. Rename the reads_file.tsv and depth_file.tsv so that the Sample column from the reads_file.tsv matches the Sample column in the depth_file.tsv. We could not automate this because different projects have unique and specific sample names :
```{bash, eval = FALSE}
head reads_file_renamed.tsv
```

```{r table5, echo=FALSE, message=FALSE, warnings=FALSE, results='asis'}
tabl_5 <- "

| Sample      | Reads          | 
|---------------|:-------------|
| SF_Jul11_sed_USGS_4_1     | 135081488   |
| SF_Jan12_sed_USGS_4_1     | 140529442   |
"
cat(tabl_5) # output the table in a format good for HTML/PDF/docx conversion
```

```{bash, eval = FALSE}
head depth_file_renamed.tsv
```

```{r table6, echo=FALSE, message=FALSE, warnings=FALSE, results='asis'}
tabl_6 <- "

| Original_Contig_Name        | contigLen           | totalAvgDepth        | Sample           | Depth       | 
|---------------|:-------------|:---------------|:-------------|:---------------:| 
|S4C_scaffold_7529_c1 | 6042.0  | 23.1232  | SF_Jul11_USGS_4_1   | 23.1232  |
|S4C_scaffold_8778_c1 | 5493.0  | 9.78027  | SF_Jul11_USGS_4_1   | 9.78027  |
"
cat(tabl_6) # output the table in a format good for HTML/PDF/docx conversion
```

### The following examples show how you could potentially rename these files:

To rename the Sample column in the reads_file.tsv, example
```{bash, eval = FALSE}
sed 's/\/fastq\//\t/g' reads_file.tsv | cut -f2,3 | sed 's/_filter-METAGENOME.fastq.gz//g' | sed 's/Reads/Sample\tReads/g' | sed 's/sed_//g' > reads_file_renamed.tsv
```

To rename the Sample column in the depth_file.tsv, example:

1. View unique names that need to be renamed
```{bash, eval = FALSE}
cut -f3 depth_file.tsv | sort | uniq
```

2. Rename
```{bash, eval = FALSE}
sed 's/-sorted.bam//g' depth_file.tsv | sed 's/\.fasta.2000kb//g' > depth_file_renamed.tsv
```

3. Make sure the names in the Sample colum of depth_file_renamed.tsv match the names of Sample in reads_file_renamed.tsv
```{bash, eval = FALSE}
cut -f3 depth_file_renamed.tsv | sort | uniq

cut -f1 reads_file_renamed.tsv | sort
```

## Calculate relative bin abundance

1. The `bin_abundance.py` script then takes in a list of input files (processed from by running the `files_prep.py` script) that need to be created beforehand. Here are the descriptions of each of the input files:
    + make sure headers of input files are tab separated
    + make sure are giving absolute paths to files (no symbolic links)
    + make sure readablenum is large enough to convert all RelativeAbundance values to greater than 1

```{bash, eval=FALSE}
usage: bin_abundance.py [-h] -r READS -m MAPPING -d DEPTH -s BINSIZE
                        [-n READABLENUM]

optional arguments:
  -h, --help            show this help message and exit
  -r READS, --reads READS
                        Total number of reads. Rows are samples column are
                        reads
  -m MAPPING, --mapping MAPPING
                        Tabular file containing Original_Contig_Name Bin Sample
  -d DEPTH, --depth DEPTH
                        Tabular file depth info Original_Contig_Name contigLen
                        Saple_Depth Depth
  -s BINSIZE, --binsize BINSIZE
                        Tabular file with Bin and corresponding Genome size
                        (bp) as columns
  -n READABLENUM, --readablenum READABLENUM
                        A large number to multiply the relative abundances so
                        that it is human readable.
```

Here is an example of how to run the script (using example data provided in the `data` folder): 

```{bash, eval=FALSE}
python3 bin_abundance.py -r ../../data/example_input_files/bin_abundance/reads_file_renamed.tsv -m ../../data/example_input_files/bin_abundance/mapping_file.tsv -d ../../data/example_input_files/bin_abundance/depth_file_renamed.tsv -s ../../data/example_input_files/bin_abundance/binsize_file.tsv -n 1000000
```

2. The output will be a file saved in the `output` directory. This file can then be used in the next script: `bin_abundance_viz.py` to visualize the data.

## Visualize relative bin abundance

*ALL input files for this step should be in either tsv or csv format.*

1. With the `bin_abundance_viz.py` script, the files outputted from `bin_abundance.py` can be used to visualize the relative bin abundances in addition to a files mapping each bin to their respective taxa (see "Mapping taxa" for more information) and each sample to their respective site (see "Mapping sample to site" for more information). These are the arguments needed for the visualization script:

```{bash, eval=FALSE}
usage: bin_abundance_viz.py [-h] -b BIN_ABUNDANCE -t TAXONOMY_INFO -s
                            SAMPLE2SITE [-p PERCENT] [-w WIDTH] [-l HEIGHT]
                            [-d DPI] [-o OUT_FIG] [-c TAXA_COLOR]

optional arguments:
  -h, --help            show this help message and exit
  -b BIN_ABUNDANCE, --bin_abundance BIN_ABUNDANCE
                        Input tsv file outputted from the bin_abundance.py
                        script with file extension.
  -t TAXONOMY_INFO, --taxonomy_info TAXONOMY_INFO
                        Input tsv or csv file mapping taxonomy to each bin
                        with file extension.
  -s SAMPLE2SITE, --sample2site SAMPLE2SITE
                      Input tsv or csv file containing each sample to the
                      site it was taken from with file extension [""].
  -p PERCENT, --percent PERCENT
                        Percent of highest sample in each bin [10].
  -w WIDTH, --width WIDTH
                        Width of outputted clustermap figure [4].
  -l HEIGHT, --height HEIGHT
                        Height of outputted clustermap figure [5].
  -d DPI, --dpi DPI     Resolution for output figure file [300].
  -o OUT_FIG, --out_fig OUT_FIG
                        Stores the figure in the specified file path and
                        format [test.png].
  -c TAXA_COLOR, --taxa_color TAXA_COLOR
                        Input tsv or csv file containing the RGB color code
                        for each taxa with file extension [""].
```  

Example of bin_abundance file:

```{bash, eval=FALSE}
head bin_abundance.tsv
```

```{r table7, echo=FALSE, message=FALSE, warnings=FALSE, results='asis'}
tabl7 <- "
| Bin             | Sample  | Sampling_Site     | RelativeAbundance       | RelativeAbundanceReadable   |
|---------------|:---------------|:---------------|:---------------|:---------------:|
| 13_May_SF_Bin93  | SF_May12_USGS_13   | S1B   | 6.6878693485613E-08     | 6.687869348561300   |
| 8_1_Oct_SF_Bin28 | SF_Oct11_USGS_8_1  | S5D   | 5.76751058810138E-08    | 5.767510588101380   |
| 4_1_Jan_SF_Bin44 | SF_Jan12_USGS_4_1  | S4A   | 1.52873009442108E-07    | 15.28730094421080   |
| 24_July_SF_Bin15 | SF_Jul11_USGS_24   | S3C   | 9.88355407641773E-08    | 9.883554076417730   |
"

cat(tabl7)
```

Example of taxonomy_info file:

```{bash, eval=FALSE}
head taxonomy_info.tsv
```

```{r table8, echo=FALSE, message=FALSE, warnings=FALSE, results='asis'}
tabl8 <- "
| Bin             | Taxa    |
|---------------|:---------------:|
| 13_Jan_SF_Bin36 | Pacearchaeota                   |
| 13_Jan_SF_Bin40 | Thaumarchaeota_Nitrosopumilus   |
| 13_Jan_SF_Bin52 | Thaumarchaeota_Nitrosopumilus   |
| 13_Jan_SF_Bin61 | Thermoplasmata                  |
"

cat(tabl8)
```

Example of sample2site file - the Sample names should match the names in the Sample column of `reads_file.tsv`; the Site names are the physical location's name/code that each sample was taken from:

```{bash, eval=FALSE}
head sample2site.tsv
```

```{r table9, echo=FALSE, message=FALSE, warnings=FALSE, results='asis'}
tabl9 <- "
| Sample             | Site   |
|---------------|:---------------:|
| SF_Jan12_USGS_1    | 21   |
| SF_Jan12_USGS_2    | 24   |
| SF_Jul11_USGS_3    | 13   |
| SF_Jul11_USGS_1    | 21   |
"

cat(tabl9)
```

Example of taxa_color file:

```{bash, eval=FALSE}
head taxa_color.tsv
```

```{r table10, echo=FALSE, message=FALSE, warnings=FALSE, results='asis'}
tabl10 <- "
| #a3c1ad
| #ede4d9
| #426ca6
| #ffdab9
| #fff5e5
| #ffefd5
"

cat(tabl10)
```

Here is an example of how to run this script:

```{bash, eval=FALSE}
python3 bin_abundance_viz.py -b ../output/bin_abundance_output.tsv -t ~/tax2bin.tsv -p 10 -w 8 -l 5 -d 500 -o figure1.jpg -c ~/taxa_color.tsv
```

2. Outputs will be saved in the `output` directory containing the user-specified name.

# Running metabolism

*NOTE: The only metabolic pathways that are available to analyze are in the KEGG, COG, PFAM, and EC_NUMBER databases.*

## Map metabolic pathways to the scaffold containing them and get the counts of each pathway per bin

1. One requirement must be met before running the `metabolic_profile.py` script, which is that the IMG annotation files must be combined into one (see "Creating IMG annotation" for more information).

2. To get the pathways mapped to each scaffold and the number of times each pathway is found within a bin, run the `metabolic_profile.py` script. These are the arguments for this script:

```{bash, highlight=TRUE, eval=FALSE}
usage: metabolic_profile.py [-h] -i IMGANNO -m MAPPING -d DATABASE [-c CONSISTENCY]

optional arguments:
  -h, --help            show this help message and exit
  -i IMGANNO, --imganno IMGANNO
                        Input file or files in tsv format. Rows are genes columns are IMG annotations.
  -m MAPPING, --mapping MAPPING
                        A tsv file containing original contig name, sample, and bin columns. Created from the files_prep.py script.
  -d DATABASE, --database DATABASE
                        Database(s) of interest to merge together. Input as a list.
  -c CONSISTENCY, --consistency CONSISTENCY
                        Boolean value that determines if scaffolds not containing a value for each database should be kept. Leave blank if consistency check is not needed.
```

Example of imganno file:

```{bash, eval=FALSE}
head imganno1.tsv
```

```{r table11, echo=FALSE, message=FALSE, warnings=FALSE, results='asis'}
tabl11 <- "
| Original_Contig_Name    | IMG_Contig_Name | Locus_Tag       | IMG_Gene_ID     | Gene_Type       | Gene_Start      | Gene_Stop       | Gene_Length     | Homolog_Gene_ID   | Homolog_Taxon_ID        | Lineage_%ID     |
|---------------|:---------------|:---------------|:---------------|:---------------|:---------------|:---------------|:---------------|:---------------|:---------------|:---------------:|
| S1A_scaffold_1_c2       | Ga0224507_10000001      | Ga0224507_100000011     | Ga0224507_10000001.1    | CDS     | 2       805     | 803     | 641316135       | 641228499       | 65.06   | Archaea;Thaumarchaeota;unclass  |
| S1A_scaffold_1_c2       | Ga0224507_10000001      | Ga0224507_100000012     | Ga0224507_10000001.3    | CDS     | 943     | 1254    | 311     | 2518759324      | 2518645532      | 79.61   | Archaea;Thaumarchaeota;unclass  |
| S1A_scaffold_1_c2       | Ga0224507_10000001      | Ga0224507_100000013     | Ga0224507_10000001.5    | CDS     | 1243    | 1911    | 668     | 2518890842      | 2518645576      | 82.81   | Archaea;Thaumarchaeota;unclass  |
| S1A_scaffold_1_c2       | Ga0224507_10000001      | Ga0224507_100000014     | Ga0224507_10000001.7    | CDS     | 1945    | 3255    | 1310    | 641316132       | 641228499       | 88.97   | Archaea;Thaumarchaeota;unclass  |
| S1A_scaffold_1_c2       | Ga0224507_10000001      | Ga0224507_100000015     | Ga0224507_10000001.9    | CDS     | 3248    | 3595    | 347     | 2518890840      | 2518645576      | 72.17   | Archaea;Thaumarchaeota;unclass  |
"

cat(tabl11)
```

Example of mapping file:

```{bash, eval=FALSE}
head mapping_file.tsv
```

```{r table12, echo=FALSE, message=FALSE, warnings=FALSE, results='asis'}
tabl12 <- "
| Original_Contig_Name    | Bin     | Sampling_Site   | 
|---------------|:---------------|:---------------:|
| S1B_scaffold_1283_c1    | 13_May_SF_Bin93 | S1B     | 
| S1B_scaffold_1516_c1    | 13_May_SF_Bin93 | S1B     | 
| S1B_scaffold_1603_c1    | 13_May_SF_Bin93 | S1B     | 
| S1B_scaffold_1628_c1    | 13_May_SF_Bin93 | S1B     | 
"

cat(tabl12)
```

Here is an example of how to run this script with a single IMG annotation file:

```{bash, highlight=TRUE, eval=FALSE}
python3 metabolic_profile.py -i imgannotation.tsv -m ../../data/example_input_files/bin_abundance/mapping_file.tsv -d [KEGG, COG, PFAM, EC_NUMBER] -c True
```

Here is an example of how to run this script with multiple IMG annotation files (escape character needed for the asterisk *):

```{bash, highlight=TRUE, eval=FALSE}
python3 metabolic_profile.py -i /\*_imgannotation.tsv -m ../../data/example_input_files/bin_abundance/mapping_file.tsv -d [KEGG, COG, PFAM, EC_NUMBER] -c True
```

2. Outputs will be saved in the `output` directory containing name formats: "mapped_scaffolds.tsv" and the database name followed by "metabolic_profile.tsv". NOTE: Old files will NOT be removed as each output is uniquely named.

## Optional step: Merge databases into one file

*This step is optional if the user has multiple databases that they need to merge into one file for the next step.*

1. To merge each of the databases, run the `merge_database.py` script. These are the arguments for the script:

```{bash, eval=FALSE}
usage: merge_databases.py [-h] -p DATABASES_PATH -d DATABASE

optional arguments:
  -h, --help            show this help message and exit
  -p DATABASES_PATH, --databases_path DATABASES_PATH
                        Path for directory containing all the database files.
  -d DATABASE, --database DATABASE
                        Database(s) of interest to merge together. Input as a
                        list.
```

Here is an example of how to run this script:

```{bash,eval=FALSE}
python3 merge_database.py -p ../../output/ -d [KEGG, COG, PFAM, EC_NUMBER]
```

2. Outputs will be places in the `output` directory containing the name formats: "merged_" database name "_metabolic_profile.tsv" and "final_merged_metabolic_profile.tsv".

## Extract specific pathways to analyze

1. One requirement must be met to run the `pathway_extraction.py` script, which is that the user-defined database must be created (see "Defining metabolic database" for more information).

2. To extract and analyze specific metabolic pathways, run the `pathway_extraction.py` script. These are the arguments for the script:

```{bash,eval=FALSE}
usage: pathway_extraction.py [-h] -c CUSTOMDATA -p PATHWAY_DATABASE -d DATABASE

optional arguments:
  -h, --help            show this help message and exit
  -c CUSTOMDATA, --customdata CUSTOMDATA
                        Input file in tsv format with at least one column with the header containing the pathways of interest. The headers of the metabolic pathway must correspond with its
                        respective database (eg. KEGG, COG, PFAM, or EC_NUMBER).
  -p PATHWAY_DATABASE, --pathway_database PATHWAY_DATABASE
                        Input file in tsv format that contains the counts of each pathways per bin.
  -d DATABASE, --database DATABASE
                        Database(s) of interest to merge together. Input as a list.
```

Example of customdata file (can have other descriptive columns if needed):

```{bash,eval=FALSE}
head custom_database.tsv
```

```{r table13, echo=FALSE, message=FALSE, warnings=FALSE, results='asis'}
tabl13 <- "
| Pathway    | Subpathway   | KEGG    | COG      | PFAM     | Description   | 
|---------------|:---------------|:---------------|:---------------|:---------------|:---------------:|
| Nitrogen   | Annamox      | K20932  | COG3391  | PF18582  | K20932; hydrazine synthase subunit [EC:1.7.2.7]   | 
| Nitrogen   | Annamox      | K20933  | COG1858  |          | K20933; hydrazine synthase subunit [EC:1.7.2.7]   | 
| Nitrogen   | Annamox      | K20934  | COG0823  |          | K20934; hydrazine synthase subunit [EC:1.7.2.7]   | 
| Nitrogen   | Annamox      | K20935  |          |          | hdh; hydrazine dehydrogenase [EC:1.7.2.8]   | 
"

cat(tabl13)
```

Example of pathway_database file:

```{bash,eval=FALSE}
head final_merged_metabolic_profile.tsv
```

```{r table14, echo=FALSE, message=FALSE, warnings=FALSE, results='asis'}
tabl14 <- "
| KEGG        | 13_Jan_SF_Bin1  | 13_Jan_SF_Bin10  | 13_Jan_SF_Bin12  | 13_Jan_SF_Bin13  | 13_Jan_SF_Bin14  | 13_Jan_SF_Bin2  | 13_Jan_SF_Bin21   | 
|---------------|:---------------|:---------------|:---------------|:---------------|:---------------|:---------------|:---------------:|
| KO:K00001   | 0.0             | 0.0              | 0.0              | 0.0              | 0.0              | 0.0             | 0.0   | 
| KO:K00002   | 1.0             | 0.0              | 0.0              | 0.0              | 0.0              | 0.0             | 0.0   |   
| KO:K00003   | 2.0             | 0.0              | 1.0              | 1.0              | 0.0              | 0.0             | 1.0   |   
| KO:K00004   | 0.0             | 0.0              | 0.0              | 0.0              | 0.0              | 0.0             | 0.0   |    
| KO:K00005   | 0.0             | 0.0              | 0.0              | 0.0              | 0.0              | 0.0             | 0.0   |    
| KO:K00006   | 0.0             | 0.0              | 0.0              | 0.0              | 0.0              | 0.0             | 0.0   |    
| KO:K00008   | 0.0             | 0.0              | 0.0              | 0.0              | 0.0              | 0.0             | 0.0   | 
"

cat(tabl14)
```

Here is an example of how to run this script:

```{bash, eval=FALSE}
python3 pathway_extraction.py -c user-defined_database.tsv -p ../../output/final_merged_metabolic_profile.tsv -d [KEGG, COG, PFAM, EC_NUMBER]
```

3. Outputs will be places in the `output` directory containing the name "extracted_pathways.tsv".

# Running host-virus

## Compare phage and host metabolisms to identify commonalities

1. There are two requirements needed to run this script: 1) the "database_counts" file outputted using the `merge_databases.py` script from the "metabolism" portion of this pipeline and 2) the Vibrant output file mapping scaffolds to metabolic pathways.

2. To analyze the unique and shared metabolic pathways between the two organisms, run the `amg_hostvi.py` script. These are the arguments for the script:

```{bash,eval=FALSE}
usage: amg_hostvi.py [-h] -p PATHWAY_DATABASE -v VIBRANT_OUTOUT -d
                               DATABASE

optional arguments:
  -h, --help            show this help message and exit
  -p PATHWAY_DATABASE, --pathway_database PATHWAY_DATABASE
                        Input file in tsv format that contains the counts of
                        each pathways per bin.
  -v VIBRANT_OUTOUT, --vibrant_outout VIBRANT_OUTOUT
                        Input file from Vibrant in tsv format that contains
                        phage scaffolds mapped to metabolic pathways.
  -d DATABASE, --database DATABASE
                        Input only one of the following database names the
                        user is interested in analyzing: KEGG, COG, PFAM, or
                        EC_NUMBER.
```

Example of pathway_database file:

```{bash,eval=FALSE}
head final_merged_metabolic_profile.tsv
```

```{r table15, echo=FALSE, message=FALSE, warnings=FALSE, results='asis'}
tabl15 <- "
| KEGG        | 13_Jan_SF_Bin1  | 13_Jan_SF_Bin10  | 13_Jan_SF_Bin12  | 13_Jan_SF_Bin13  | 13_Jan_SF_Bin14  | 13_Jan_SF_Bin2  | 13_Jan_SF_Bin21   | 
|---------------|:---------------|:---------------|:---------------|:---------------|:---------------|:---------------|:---------------:|
| KO:K00001   | 0.0             | 0.0              | 0.0              | 0.0              | 0.0              | 0.0             | 0.0   | 
| KO:K00002   | 1.0             | 0.0              | 0.0              | 0.0              | 0.0              | 0.0             | 0.0   |   
| KO:K00003   | 2.0             | 0.0              | 1.0              | 1.0              | 0.0              | 0.0             | 1.0   |   
| KO:K00004   | 0.0             | 0.0              | 0.0              | 0.0              | 0.0              | 0.0             | 0.0   |    
| KO:K00005   | 0.0             | 0.0              | 0.0              | 0.0              | 0.0              | 0.0             | 0.0   |    
| KO:K00006   | 0.0             | 0.0              | 0.0              | 0.0              | 0.0              | 0.0             | 0.0   |    
| KO:K00008   | 0.0             | 0.0              | 0.0              | 0.0              | 0.0              | 0.0             | 0.0   | 
"

cat(tabl15)
```

Example of vibrant_output file

```{bash,eval=FALSE}
head vibrant_output.tsv
```

```{r table16, echo=FALSE, message=FALSE, warnings=FALSE, results='asis'}
tabl16 <- "
| protein             | scaffold        | AMG KO   | AMG KO name                                            | Pfam         | Pfam name   |
|---------------|:---------------|:---------------|:---------------|:---------------|:---------------:|
| scaffold_1527_10    | scaffold_1527   | K03639   | \"moaA, CNX2; GTP 3',8-cyclase [EC:4.1.99.22]\"          | PF04055.21   | Radical SAM superfamily   |
| scaffold_1527_19    | scaffold_1527   | K01939   | \"purA, ADSS; adenylosuccinate synthase [EC:6.3.4.4]\"   | PF00709.21   | Adenylosuccinate synthetase   |
| scaffold_224_117    | scaffold_224    | K01939   | \"purA, ADSS; adenylosuccinate synthase [EC:6.3.4.4]\"   | PF00709.21   | Adenylosuccinate synthetase   |
| scaffold_224_158    | scaffold_224    | K22227   | \"ahbD; heme synthase\"                                  | PF13186.6    | Iron-sulfur cluster-binding domain   |
"

cat(tabl16)
```

Here is an example of how to run this script:

```{bash,eval=FALSE}
python3 amg_hostvi.py -p ../../output/KEGG_database_counts.tsv -v ~/vibrant_output.tsv -d KEGG
```

3. Outputs will be places in the `output` directory containing the names "phage_host_metabolism.tsv" and "phage_host_mapping.tsv".

# Additional commands

## Download fastqs

Follow the download instructions found on [IMG](https://genome.jgi.doe.gov/portal/help/download.jsf#/api).

## Mapping bins to sample site

The user must manually map each bin to its respective sample location.

## Download and annotate genomes using IMG

Annotate metagenomes at the [IMG JGI platafform](https://img.jgi.doe.gov/cgi-bin/mer/main.cgi) 

After completion, you will download a directory of each of your metagenomic samples containing the following subdirectories: 

1. **IMG Data**: Contanining the following files 

```
207502.assembled.COG
207502.assembled.crisprs
207502.assembled.EC
207502.assembled.faa
207502.assembled.fna
207502.assembled.gff
207502.assembled.KO
207502.assembled.last.blasttab
207502.assembled.names_map
207502.assembled.Pfam.blout
207502.assembled.phylodist
207502.assembled.product_names
207502.assembled.unsorted.Pfam.hmmout
3300033388
```

2. **Metagenome Report Tables**: containing the following files 

```
Table_6_-_207017.annotation_parameters.txt
Table_7_-_3300033141.metagenome_properties.txt
Table_8_-_3300033141.taxonomic_composition.txt
Table_9_-_3300033141.functional_diversity.txt
```

## Create initial depth files

After you have annotated and downloaded your genomes using IMG's JGI platform, there are two possible methods to create the initial depth txt files. The following two methods use BWA and JGI, respectively.

### BWA method

Here is an example of how to create the depth txt files using BWA:

1. Prep samples one at time for mapping.

```{bash, eval=FALSE}
#Copy fq files from yggshare to folder
cp 4998_22_23_H2_S10_fastp_bb.fastq.gz /home/keappler/07_Lab_Protocols/Binning/Guaymas_2018_Samples/mapping_bwa/
  
#Run mdsum
for i in *.gz ; do md5sum $i > $i.md5sum ; done
```

2. Run script example for one sample below.

```{bash, eval=FALSE}
#STARTED 10:39 4/5/2021
#!/bin/bash
fq=/home/keappler/07_Lab_Protocols/Binning/Guaymas_2018_Samples/mapping_bwa/4998_22_23_H2_S10_fastp_bb.fastq.gz
assembly=/home/keappler/07_Lab_Protocols/Binning/Guaymas_2018_Samples/mapping_bwa/2.5kb/renamed/2.5kb_4998_22_23_H2_S10.headers.removed.fa

#Index contigs >=2.5kb assembly subset created in previous steps

bwa index $assembly 
echo "Done index" 

# Align contigs >=2.5kb against concatenated, interleaved, and trimmed R1 and R2
bwa mem -t 70 $assembly -p $fq > 4998_22_23_H2_S10.sam
echo "Done mapping" 

#samtools create bam files- needed for binning
samtools view -bS -F 4 -@10 4998_22_23_H2_S10.sam -o 4998_22_23_H2_S10.bam
echo "Done bam" 

#samtools create sorted bam files- needed for jgi_summarize_bam_contig_depths
samtools sort -@10 4998_22_23_H2_S10.bam -o 4998_22_23_H2_S10_sorted.bam
echo "Done sorted bam"

#The file to put the contig by bam depth matrix > depth.txt
#Do not include variance from mean depth along the contig
jgi_summarize_bam_contig_depths  --outputDepth depth.txt 4998_22_23_H2_S10_sorted.bam --noIntraDepthVariance
echo "Done depth file"
```

3. Run md5sum and delete .sam file (large) before starting next sample.

```{bash, eval=FALSE}
for i in *.sam ; do md5sum $i > $i.md5sum ; done
```

4. Repeat process for each sample.

### JGI method

The script, `jgi_summarize_bam_contig_depths.cpp`, must be used to create the depth file needed in `files_prep.py` and can be found [here](https://bitbucket.org/berkeleylab/metabat/src/master/) along with its usage notes.

Here is an example of how to create the depth txt files using JGI:

1. Index bin.

```{bash, eval=FALSE}
bwa index 13_Jan_SF_Bin5.fna
```

2. BWA MEM to obtain SAM files.

```{bash, eval=FALSE}
for sample in `cat names.txt`; do bwa mem -t 30 13_Jan_SF_Bin5.fna -p /home/mlangwig/Working_Dir/SF_Bay/fastq/$sample.fastq.gz > $sample.sam; done
```

3. Samtools view to convert SAM to BAM.

```{bash, eval=FALSE}
for sample in `cat names.txt`; do samtools view -bS -F 4 -@10  $sample.sam -o $sample.bam; done
```

4. Obtain sorted bams.

```{bash, eval=FALSE}
for sample in `cat names.txt`; do samtools sort -@10 $sample.bam -o $sample-sorted.bam; done
```

5. Use sorted bams to get Cov_ files for mmgenome2.

```{bash, eval=FALSE}
jgi_summarize_bam_contig_depths --outputDepth depth.txt *sorted.bam --noIntraDepthVariance
```

6. Repeat process for each sample.

## Mapping taxa

One way to map each the taxa to each bin is to use the GTDBTK tool found here: https://github.com/Ecogenomics/GTDBTk.

## Creating IMG annotation

First, log on to the IMG server found [here](https://genome.jgi.doe.gov/portal/help/download.jsf#/api). Then, download and unzip the tar.gz file. Once you have the IMG annotation files, run the `consolidateJGIdata.pl` script to consolidate the data obtained from JGI IMG into one file. The script can be found [here](https://github.com/Geo-omics/scripts).

## Defining metabolic database

When creating a user-defined database, there are multiple format requirements. The metabolic databases of interest should contain one or multiple headers (KEGG, COG, PFAM, EC_NUMBER) depending on which database the user wants to analyze. All other auxiliary columns will be mapped to the bins and added to the final outputted file. Lastly, in each of the database columns, please format each pathway name as it is in the IMG annotation file:

```{bash,eval=FALSE}
KEGG - KO:K12345
COG - COG12345
PFAM - pfam12345
EC_NUMBER - EC:EC12345
```
## Creating Vibrant mapping

I need help with this.

# Support and Development

This pipeline has been developed by the Baker lab at UT Austin. If there are any questions, comments, or concerns, please email us at either sahilbshah@berkeley.edu or valdeanda@utexas.edu.
 
# Cite 
If you find this software useful please cite us as: 

Coming soon!

